plyranges: a grammar for transforming genomics data

### 2018-04-09 paper.Rmd
shuffling the intro around, adding some of Michael's changes.

High-throughput genomics promises to unlock new disease therapies,
reveal basic biology, and solve other problems that plague the human
condition. To deliver on those promises, scientists must derive a
stream of knowledge from a deluge of data. Genomic data is challenging
in both scale and complexity. Innovations in sequencing technology
often outstrip our capacity to process the output. Beyond their common
association with genomic coordinates, genomic data are heterogeneous,
consisting of raw sequence read alignments, genomic feature
annotations like genes and exons, and summaries like coverage vectors,
ChIP-seq peak calls, variant calls, and per-feature read
counts. Genomic scientists need software tools to wrangle the
different types of data, process the data at scale, test hypotheses,
and generate new ones, all while focusing on the biology, not the
computation. For the tool developer, the challenge is to define ways
to model and operate on the data that align with the mental model of
scientists, and to provide an implementation that scales with their
ambition.

Several domain specific languages (DSLs) enable scientists to process
and reason about heterogenous genomics data by expresssing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED.  An example of the former is the Genome
Query Language (GQL) and its distributed implementation GenAp which
use an SQL-like syntax for fast retrieval of information of
unprocessed sequencing data @Kozanitis2014-va; @Kozanitis2016-bm.
Similarly, the Genometric Query Language (GMQL) implements a
relational algebra for combining genomic datasets @Kaitoua2017-pw. The
command line application BEDtools defines an extensive algebra for
performing arithmetic between two or more sets of genomic regions
stored as files @Quinlan2010-gc; @Dale2011-js. All of the
aformentioned DSLs are designed to be evaluated either at the command
line or embedded in scripts for batch processing. They exist in a
sparse ecosystem, mostly consisting of UNIX and database tools that
lack biological semantics and operate at the level of files and
database tables.

The Bioconductor packages `IRanges` and `GenomicRanges`
(@Lawrence2013-wg; @Huber2015-ei) define a DSL for analysing genomics
data with R, an interactive data analysis environment that encourages
reproducibility and provides high-level abstractions for manipulating,
modeling and plotting data, through state of the art methods in
statistical computing. The packages define object-oriented (OO)
abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. The Bioconductor
infrastructure models the genomic data and operations from the
perspective of the power user, one who understands and wants to take
advantage of the subtle differences in data types. This design has
enabled the development of sophisticated tools, as evidenced by the
hundreds of packages depending the framework. Unfortunately, the
myriad of data structures have overlapping purposes and important but
obscure differences in behavior that often confuse the typical end
user.

Recently, there has been a concerted, community effort to standardize
R data structures and workflows around the notion of tidy data
@Wickham2014-jc. A tidy dataset is defined to have a tabular structure
that has observations as rows and columns as variables. Each tidy
dataset represents measurements from a single observational unit.  The
tidy data pattern is useful because it allows us to see how the data
relates to the design of an experiment and the variables measured.
The `dplyr` pacakge [@Wickham2017-dplyr] defines an API that maps
notions from the general relational algebra to operations on tidy
data. It expresses each operation as a cohesive, endomorphic
verb. Taken together these features enable a user to write human
readable analysis workflows.

@@

# Stuff pushed out of the intro (needs a home)

The most general Bioconductor data structure, GRanges, is capable of
representing all types of genomic data at a semantic level that
roughly matches the intuition of most users.  A GRanges is essentially
a table, with columns for the chromosome, stand and end coordinates,
and the strand, along with an arbitrary set of additional columns,
consisting of measurements or metadata specific to the data type or
experiment.  For example, a gene can be represented by its chromosomal
coordinates, along with its identifier and the identifiers of its
exons; or an RNA-seq experiment may be represented as sets of genes
with a matching count column.

GRanges follows the tidy data pattern: it is a rectangular table
corresponding to a single biological context. Each row contains a
single observation and each column is a variable about that
observation.

Cohesion means that each function performs a singular task and has no
side-effects, while endomorphism means a function call returns the
same type of object as its input.

@@

>>>>

High-throughput genomics promises to unlock new disease therapies,
and strengthen our knowledge of basic biology. To deliver on those promises, 
scientists must derive a stream of knowledge from a deluge of data. Genomic 
data is challenging in both scale and complexity. Innovations in sequencing 
technology often outstrip our capacity to process the output. Beyond their common
association with genomic coordinates, genomic data are heterogeneous,
consisting of raw sequence read alignments, genomic feature
annotations like genes and exons, and summaries like coverage vectors,
ChIP-seq peak calls, variant calls, and per-feature read
counts. Genomic scientists need software tools to wrangle the
different types of data, process the data at scale, test hypotheses,
and generate new ones, all while focusing on the biology, not the
computation. For the tool developer, the challenge is to define ways
to model and operate on the data that align with the mental model of
scientists, and to provide an implementation that scales with their
ambition.

Several domain specific languages (DSLs) enable scientists to process and reason
about heterogenous genomics data heterogenous genomics data by expresssing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED.
An example of the former is the Genome Query Language (GQL) and its distributed 
implementation GenAp which use an SQL-like syntax for fast retrieval of 
information of unprocessed sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. 
Similarly, the Genometric Query Language (GMQL) implements a 
relational algebra for combining genomic datasets @Kaitoua2017-pw.  
The command line application BEDtools develops an extensive algebra for 
performing arithmetic between two or more sets of genomic regions 
@Quinlan2010-gc; @Dale2011-js. All of the
aformentioned DSLs are designed to be evaluated either at the command
line or embedded in scripts for batch processing. They exist in a
sparse ecosystem, mostly consisting of UNIX and database tools that
lack biological semantics and operate at the level of files and
database tables.

The Bioconductor/R packages `IRanges` and `GenomicRanges` 
[@Lawrence2013-wg; @Huber2015-ei] define a DSL for analysing genomics data with 
R, an interactive data analysis environment that encourages
reproducibility and provides high-level abstractions for manipulating,
modeling and plotting data, through state of the art methods in
statistical computing. The packages define object-oriented (OO)
abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. The Bioconductor
infrastructure models the genomic data and operations from the
perspective of the power user, one who understands and wants to take
advantage of the subtle differences in data types. This design has
enabled the development of sophisticated tools, as evidenced by the
hundreds of packages depending the framework. Unfortunately, the
myriad of data structures have overlapping purposes and important but
obscure differences in behavior that often confuse the typical end
user.

Recently, there has been a concerted, community effort to standardize
R data structures and workflows around the notion of tidy data
@Wickham2014-jc. A tidy dataset is defined to be to a tabular
data structure that has observations as rows and columns as variables, and each
tidy dataset represents measurements from a single observational unit.
The tidy data pattern is useful because it allows 
us to see how the data relates to the design of an experiment
and the variables measured. The genomic interval data structure follows 
this pattern: it is a rectangular table corresponding to a single biological 
context. Each row contains a single observation and each column is a variable about
that observation.

@@

@@

via
 This interface
provides abstractions for representing genomics data and enables interoperability
by allowing users and developers to use these abstractions in their own code
and packages. However, due to the large number of abstractions there are costs
involved. Due to the OO interface each abstraction will behave slightly 
differently for any given generic method. There is a cognitive load on the user
to understand the appropriate context for each abstraction. An alternative approach
is to recognise that most genomic data analysis tasks can be performed with 
data structures that naturally represent genomic intervals. That is, genomics
data can be represented as sets of integer pairs, where each pair has a start
and an end coordianate, along with strings representing strand and chromosome 
location. Measurements or metadata can be included as additional annotations
on each pair.  As an example. a gene can be represented with its coordinates, along with its 
identifier and the identifiers of its exons; or an RNA-seq experiment may be
represented as sets of genes with a matching count column.

The tidy data abstraction enables the construction of fluent application programming interfaces (APIs) [@Fowler2010-zd] that are optimised for readability. An example of a 
fluent interface built on the tidy data pattern is the R pacakge `dplyr` [@Wickham2017-dplyr]. 
The package defines a consistent syntax and style for transforming tabular data sets via
functions that are named as verbs. Each function performs an action on the data
set and is cohesive and endomorphic. Cohesion means that each function performs
a singular task and has no side-effects, while endomorphism means a function
call returns the same type of object as its input. Taken together these
features enable a user to write human readable analysis workflows.
